---
title: "DHS-GPS-weather"
author: "Jade TK Chui"
date: "2025-02-19"
output: html_document
---

packages
```{r}
library(haven)
library(dplyr)
library(stringr)
library(lubridate)
library(zoo)
library(tidyr)
library(sf)
library(tidyverse)
library(terra)
library(plyr)
library(data.table)
```

extract DHS SSA GPS data
```{r}
pathGPS = "E:/Data/DHS_GPS/SubSaharanAfrica"

zipGPS <- list.files(path = pathGPS, 
                   pattern = '*.zip', ignore.case = TRUE, full.names = TRUE)

tpathGPS = "E:/Jade/DHS-heat/temp/GPSfile"

GPS = list()
for (i in 1:length(zipGPS))
{GPSpath <- unzip(zipGPS[i], exdir = tpathGPS)
GPS[[i]] <- GPSpath}

GPSlist = list()
for (i in 1:length(zipGPS))
{list <- unzip(zipGPS[i], list = TRUE)[,1]
lists <- list[str_detect(list, '.shp$')]
GPSlist[i] <- lists}

GPSpath <- paste0(tpathGPS,'/', GPSlist)

GPS.SSA <- ldply(GPSpath, read_sf)

gps <- GPS.SSA %>% select(DHSID, DHSCC, DHSCLUST, DHSYEAR, LATNUM, LONGNUM) %>%
  dplyr::rename(cc = DHSCC, clust = DHSCLUST, int.y = DHSYEAR, lat = LATNUM, lon = LONGNUM) %>%
  distinct(DHSID, .keep_all = T) %>%
  drop_na(DHSID, lat, lon) %>%
  filter(!lat == lon) %>%
  mutate(lon = if_else(lon<0, lon+360, lon))
```

match monthly 2m temperature with GPS data
```{r}
ym <- seq(as.Date('2000-01-01'), as.Date('2020-12-31'), by = 'month')
colname <- as.list(as.character(ym))

r1 <- lapply('E:/Data/GLO_Weather_2000-2020/t2m.nc', rast)
s1 <- do.call(c, r1)
s1

e1 <- terra::extract(s1, vect(gps[,5:6]))
e1 <- as.data.frame(e1)

colnames(e1) <- c('ID',colname)
e1 <- e1 %>%
  mutate(DHSID = gps$DHSID)

t2m <- e1 %>%
  select(-ID) %>%
  pivot_longer(!DHSID, names_to = "ym", values_to = "t2m") %>% 
  drop_na()

rm(r1, s1, e1)
```

match monthly total precipitation with GPS data
```{r}
ym <- seq(as.Date('2000-01-01'), as.Date('2020-12-31'), by = 'month')
colname <- as.list(as.character(ym))

r1 <- lapply('E:/Data/GLO_Weather_2000-2020/tp.nc', rast)
s1 <- do.call(c, r1)
s1

e1 <- terra::extract(s1, vect(gps[,5:6]))
e1 <- as.data.frame(e1)

colnames(e1) <- c('ID',colname)
e1 <- e1 %>%
  mutate(DHSID = gps$DHSID)

tp <- e1 %>%
  select(-ID) %>%
  pivot_longer(!DHSID, names_to = "ym", values_to = "tp") %>% 
  drop_na()

rm(r1, s1, e1)
```

match monthly relative humidity with GPS data
```{r}
ym <- seq(as.Date('2000-01-01'), as.Date('2020-12-31'), by = 'month')
colname <- as.list(as.character(ym))

r1 <- lapply('E:/Data/GLO_Weather_2000-2020/rh.nc', rast)
s1 <- do.call(c, r1)
s1

e1 <- terra::extract(s1, vect(gps[,5:6]))
e1 <- as.data.frame(e1)

colnames(e1) <- c('ID',colname)
e1 <- e1 %>%
  mutate(DHSID = gps$DHSID)

rh <- e1 %>%
  select(-ID) %>%
  pivot_longer(!DHSID, names_to = "ym", values_to = "rh") %>% 
  drop_na()

rm(r1, s1, e1)
```

match monthly PM2.5 data with GPS data
```{r}
ym <- seq(as.Date('2003-01-01'), as.Date('2022-12-01'), by = 'month')
colname <- as.list(as.character(ym))

r1 <- lapply('E:/Data/GLO_AirPollution_2003-2022/pm25.nc', rast)
s1 <- do.call(c, r1)
s1

e1 <- terra::extract(s1, vect(gps[,5:6]))
e1 <- as.data.frame(e1)

colnames(e1) <- c('ID',colname)
e1 <- e1 %>%
  mutate(DHSID = gps$DHSID)

pm25 <- e1 %>%
  select(-ID) %>%
  pivot_longer(!DHSID, names_to = "ym", values_to = "pm25") %>% 
  drop_na()

rm(r1, s1, e1)
```

match monthly PM10 data with GPS data
```{r}
ym <- seq(as.Date('2003-01-01'), as.Date('2022-12-01'), by = 'month')
colname <- as.list(as.character(ym))

r1 <- lapply('E:/Data/GLO_AirPollution_2003-2022/pm10.nc', rast)
s1 <- do.call(c, r1)
s1

e1 <- terra::extract(s1, vect(gps[,5:6]))
e1 <- as.data.frame(e1)

colnames(e1) <- c('ID',colname)
e1 <- e1 %>%
  mutate(DHSID = gps$DHSID)

pm10 <- e1 %>%
  select(-ID) %>%
  pivot_longer(!DHSID, names_to = "ym", values_to = "pm10") %>% 
  drop_na()

rm(r1, s1, e1)
```

match monthly CO data with GPS data
```{r}
ym <- seq(as.Date('2003-01-01'), as.Date('2022-12-01'), by = 'month')
colname <- as.list(as.character(ym))

r1 <- lapply('E:/Data/GLO_AirPollution_2003-2022/co.nc', rast)
s1 <- do.call(c, r1)
s1

e1 <- terra::extract(s1, vect(gps[,5:6]))
e1 <- as.data.frame(e1)

colnames(e1) <- c('ID',colname)
e1 <- e1 %>%
  mutate(DHSID = gps$DHSID)

co <- e1 %>%
  select(-ID) %>%
  pivot_longer(!DHSID, names_to = "ym", values_to = "co") %>% 
  drop_na()

rm(r1, s1, e1)
```

match monthly SO2 data with GPS data
```{r}
ym <- seq(as.Date('2003-01-01'), as.Date('2022-12-01'), by = 'month')
colname <- as.list(as.character(ym))

r1 <- lapply('E:/Data/GLO_AirPollution_2003-2022/so2.nc', rast)
s1 <- do.call(c, r1)
s1

e1 <- terra::extract(s1, vect(gps[,5:6]))
e1 <- as.data.frame(e1)

colnames(e1) <- c('ID',colname)
e1 <- e1 %>%
  mutate(DHSID = gps$DHSID)

so2 <- e1 %>%
  select(-ID) %>%
  pivot_longer(!DHSID, names_to = "ym", values_to = "so2") %>% 
  drop_na()

rm(r1, s1, e1)
```

match monthly NO2 data with GPS data
```{r}
ym <- seq(as.Date('2003-01-01'), as.Date('2022-12-01'), by = 'month')
colname <- as.list(as.character(ym))

r1 <- lapply('E:/Data/GLO_AirPollution_2003-2022/no2.nc', rast)
s1 <- do.call(c, r1)
s1

e1 <- terra::extract(s1, vect(gps[,5:6]))
e1 <- as.data.frame(e1)

colnames(e1) <- c('ID',colname)
e1 <- e1 %>%
  mutate(DHSID = gps$DHSID)

no2 <- e1 %>%
  select(-ID) %>%
  pivot_longer(!DHSID, names_to = "ym", values_to = "no2") %>% 
  drop_na()

rm(r1, s1, e1)
```

match daily temperature data  with GPS data
```{r}
ncF <- list.files(path = "E:/Jade/DHS-heat/temp/t2m.daily", 
                   pattern = ".nc$", ignore.case = TRUE, full.names = TRUE)

nclist <- list()
for (nc in ncF) {
  year <- as.numeric(sub(".*(\\d{4})\\.nc", "\\1", nc))
  
  start_date <- as.Date(paste0(year, "-01-01"))
  end_date <- as.Date(paste0(year, "-12-31"))
  ymd <- seq(start_date, end_date, by = 'day')
  colname <- as.list(as.character(ymd))
  
  r1 <- lapply(nc, rast)
  s1 <- do.call(c, r1)
  
  e1 <- terra::extract(s1, vect(gps[,5:6]))
  e1 <- as.data.frame(e1)
  
  colnames(e1) <- c('ID', colname)
  e1 <- e1 %>%
    mutate(DHSID = gps$DHSID)
  
  t2md <- e1 %>%
    select(-ID) %>%
    pivot_longer(!DHSID, names_to = "ymd", values_to = "t2md") %>% 
    drop_na()
  
  nclist[[nc]] <- t2md
}

t2md <- bind_rows(nclist)
rm(nclist, r1, e1, s1)
```

calculate monthly heat days
```{r}
thresholds <- c(97, 95, 92, 90)
days <- as.numeric(as.Date("2020-12-31") - as.Date("1999-12-31"))
setDT(t2md)  

hwdlist <- list()

for (threshold in thresholds) {
  
  n <- as.integer((1 - threshold / 100) * days)
  
  hw <- t2md[, .SD[order(-t2md)[1:n]], 
          by = DHSID, 
          .SDcols = c("t2md", "ymd")]
  
  setorder(hw, DHSID, ymd)
  
  hwd <- hw[,.(hwd = .N), 
            by = .(DHSID,
                   year = as.integer(substr(ymd, 1, 4)), 
                   month = as.integer(substr(ymd, 6, 7)))
  ][, th := threshold]
  
  hwdlist[[as.character(threshold)]] <- hwd
}

hwd <- rbindlist(hwdlist)
rm(hwdlist, hw, t2md)
```

calculate heat days
```{r}
thresholds <- c(97, 95, 92, 90)
days <- as.numeric(as.Date("2020-12-31") - as.Date("1999-12-31"))
setDT(t2md)  

hwlist <- list()
for (threshold in thresholds) {
  n <- as.integer((1 - threshold / 100) * days)
  
  hw <- t2md[, .SD[order(-t2md)[1:n]], 
          by = DHSID, 
          .SDcols = c("t2md", "ymd")]
  
  setorder(hw, DHSID, ymd)
  
  hw <- hw[, .(DHSID, t2md, ymd, th = threshold)]
  
  hwlist[[as.character(threshold)]] <- hw
}

hw <- rbindlist(hwlist)
rm(hwlist, t2md)
```

merge all weather data
```{r}
w <- t2m %>%
  left_join(tp, by = c("DHSID", "ym")) %>%
  left_join(rh, by = c("DHSID", "ym")) %>%
  left_join(hwd %>% ungroup() %>%
              filter(th == 97) %>%
              mutate(ym = as.character(make_date(year, month, 1))) %>% 
              select(-year, -month, -th) %>%
              dplyr::rename(hwd97 = hwd), by =c("DHSID", "ym")) %>%
  left_join(hwd %>% ungroup() %>%
              filter(th == 95) %>%
              mutate(ym = as.character(make_date(year, month, 1))) %>% 
              select(-year, -month, -th) %>%
              dplyr::rename(hwd95 = hwd), by =c("DHSID", "ym")) %>%
  left_join(hwd %>% ungroup() %>%
              filter(th == 92) %>%
              mutate(ym = as.character(make_date(year, month, 1))) %>% 
              select(-year, -month, -th) %>%
              dplyr::rename(hwd92 = hwd), by =c("DHSID", "ym")) %>%
  left_join(hwd %>% ungroup() %>%
              filter(th == 90) %>%
              mutate(ym = as.character(make_date(year, month, 1))) %>% 
              select(-year, -month, -th) %>%
              dplyr::rename(hwd90 = hwd), by =c("DHSID", "ym")) %>%
  mutate(hwd97 = if_else(is.na(hwd97) == T, 0, hwd97), 
         hwd95 = if_else(is.na(hwd95) == T, 0, hwd95), 
         hwd92 = if_else(is.na(hwd92) == T, 0, hwd92), 
         hwd90 = if_else(is.na(hwd90) == T, 0, hwd90))

rm(t2m, tp, rh)
```

merge all air pollution data
```{r}
p <- co %>%
  left_join(no2, by = c("DHSID", "ym")) %>%
  left_join(pm10, by = c("DHSID", "ym")) %>%
  left_join(pm25, by = c("DHSID", "ym")) %>%
  left_join(so2, by = c("DHSID", "ym"))

rm(co, no2, pm10, pm25, so2)
```

merge all environmental data
```{r}
env <- p %>%
  full_join(w, by = c("DHSID", "ym")) %>%
  drop_na()
```

export generated environmental data matched with DHS cluster ID
```{r}
write.csv(gps, row.names = FALSE, file = 'E:/Jade/DHS-heat/data/gps.csv')
write.csv(env, row.names = FALSE, file = 'E:/Jade/DHS-heat/data/env.csv')
write.csv(hwd, row.names = FALSE, file = 'E:/Jade/DHS-heat/data/hwd.csv')
```

